NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-01-03 main.tex

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{tablefootnote}
\usepackage{xr}

@@

\title{NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all}
%Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration}

\author[1]{Anne Kiirikki}
\author[2]{...}
\author[1,*]{O. H. Samuli Ollila}
%\author[1,2,+]{Christine Author}
%\author[2,+]{Derek Author}
\affil[1]{University of Helsinki, Institute of Biotechonology, Helsinki, Finland}
\affil[2]{Affiliation, department, city, postcode, country}

@@

Cellular membranes are composed of hundreds of different types of lipid molecules which regulate membrane properties, morphology and biological functions~\cite{vanmeer08}. Membrane lipid composition can be also related to diseases, such as cancer and neurodegenerative disorders, and potential therapeutics that affect membrane compositions have been proposed~\cite{torres21}. However, membrane-containing systems are often difficult to study experimentally because the membranes are in a disordered fluid state under biological conditions and sample preparation can be arduous due to the complicated phase behaviour and rich interactions between lipids and other biomolecules. For those reasons, the detailed connections between lipid interactions and biological functions taking place in or around membranes are still poorly understood.
%why such variety of lipids is needed and how their. 
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, yet their accuracy is often compromised by the quality of model parameters, methodological inadequacy, and other artefacts~\cite{antila22b,gupta22}. 
%While MD simulations of systems with few biomolecules continue having major contributions in our understanding on biology and drug design, 
On the  other hand, the accuracy of models is becoming increasingly important as the scale (in both size and time) of the systems studied continuously increases 
%researches are progressing from the simulations of 
%now searching new avenues by building models not only 
from the level of individual molecules to models of whole cells or organelles using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit complex emergent behaviour making inaccuracies more difficult to detect and accumulation of even modest errors may have major effects the conclusions.

In contrast to experimental structural biology, where standard protocols to share and quality evaluate the resolved structures are established in the PDB databank~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and solutions for sharing data are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, cyclodextrins~\cite{mixcoha16}, and arbitrary biomolecules~\cite{bekker20}. However,
%The relevance of quality evaluation of simulation trajectories in databanks regarding technical details of simulations and accuracy of the underlying physical description of the system (force field) has become evident~\cite{tai04,meyer10,hospital20} and such 
a streamlined workflow with automatic quality evaluation~\cite{meyer10,hospital16} and programmatic access is still rare. 
%However, straightforward quality comparisons between individual simulations or force fields within these databanks remain challenging. 
Therefore, tools for automatic quality evaluation of membrane simulations or training sets for machine learning models of membrane-containing systems are not yet available.

@@

Here we present the NMRlipids databank, a community-driven open access databank with programmatic access containing atomistic resolution MD simulations of lipid bilayers. The programmatic access enables various data-driven approaches, providing new tools for researchers in a wide range of fields covering academia and industry from cell membrane biology to lipid nanoparticle formulations, data-driven computational chemistry, and machine learning. Here, we demonstrate how data-driven analysis of water anisotropic diffusion in membrane systems from the NMRlipids databank can extend the scope of MD simulations to MRI imaging and pharmacokinetics as well as how the NMRlipids databank can be utilized to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Furthermore, we perform automatic quality evaluation of membrane simulations, which guides the selection of best models for specific applications and the improvement of simulation parameters and methodology.
%The practical relevance of the NMRlipids databank is exemplified by automatic quality evaluation and ranking of large amount of MD simulation data,

While the NMRlipids databank currently contains only bilayer systems, it can readily be expanded to contain other molecules than lipids, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The combination of an overlay databank structure and a community-wide open collaboration used here has been applied successfully when building databanks in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Thus, the approach has already proven itself most useful in enabling data-driven and machine learning applications in new fields.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Programmatic access to MD simulation data of membranes composed of most common biologically abundant lipids}
NMRlipids databank is a community-driven databank containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. Nearly a thousand simulation trajectories with the total length approaching half a millisecond can be accessed programmatically using the Python application programming interface (API) or a graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}). 
%NMRlipids databank is the only databank of MD simulations with open programmatic access to its content enabling fully flexible analyses of its content. 
Following the FAIR principles~\cite{wilkinson16} and NMRlipids project protocol, not only the content and computer programs related to the databank, but also the entire construction process of the databank, are open access~\cite{botan15}. 
%The NMRlipids databank will be always open for data and other contributions.

Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. The distribution of available simulations containing a specific lipid is shown in Fig.~\ref{fig:overlay}C. This roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. The force fields used in the simulations cover all the essential parameters commonly used in lipid simulations, ranging from highly popular CHARMM36 parameters to united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.
%Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

The efficient upcycling of large MD simulation trajectories in the NMRlipids databank is enabled by the overlay structure illustrated in Fig.~\ref{fig:overlay}A. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability and permanent links to the data (e.g., digital object identifiers (DOIs)). The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for the lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C-H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored is the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}), and results from analyses described in sections below. A more detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\begin{figure}
    \centering
    \includegraphics[width = 180mm]{Figures/overlay2.pdf}
    \caption{a) Structure of an overlay databank. 
    %Raw data is available from a publicly accessible repository (layer 1).
    %The core of the databank contains information on the location of the raw data and indexing of molecular %compositions and simulation details described with universal naming conventions (layer 2).
    %The content of databank can be viewed and analysed using external programs that automatically read the information and data from layers 1 and 2 (layer 3). Examples of such applications are the GitHub repository generating the results presented in this work and the graphical access to the databank.
    More detailed structure of the layer 2 in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total lenght of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or less simulations ('others') are listed in the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    }
    \label{fig:overlay}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding the accuracy of used force field parameters is crucial when assessing the reliability and significance MD simulation results. However, the lack of universal quality measures for MD simulations and the complicated landscape of force field quality for membranes hamper the applicability such simulations, resulting in substantial controversies~\cite{antila22b}. For example, previous quality evaluations of membrane simulations have concluded that CHARMM36 parameters give the best description for lipid headgroup conformational ensembles~\cite{bacle21}, GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}, and that OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%, thereby being a major obstacle in many applications of membrane MD simulations. 
To enable the navigation in this complex force field milieu, we here define quantitative quality measures that can be used to rapidly find the best available parameters for a specific application or to guide the development of improved force fields. For detailed definitions of quality measures, see the supplementary information.

\begin{figure}[!t]
    \centering
    \includegraphics[width=180mm]{Figures/quality3.pdf}
    \caption{ A) Best simulations ranked based on overall order parameter quality.
    B) Best simulations ranked based on the overall order parameter quality of POPE lipid. 
    C)-E) Evaluation against experimental data exemplified for a simulation with the best overall order parameter quality (C), the best quality for POPE lipid (D), and the headgroup quality for POPE (E).
    F) Scatter plots and Pearson correlation coefficients for the membrane area per lipid, thickness, first minima of x-ray scattering form factor and average order parameter of the {\it sn}-1 acyl chain extracted from the NMRlipids databank. 
    All correlation coefficients have p-value below 0.001. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

Ranking of all simulations with the highest scores based on evaluation against C-H bond order parameters from NMR experiments is shown in Fig.~\ref{fig:quality} A and ranking of best POPE lipid bilayer simulations in Fig.~\ref{fig:quality} B. Direct comparisons with experiments for selected simulations are shown in Figs.~\ref{fig:quality} C-D. C-H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16} while also correlating well with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality} F. This is complemented by comparing the absolute values of X-ray scattering form factors between simulations and experiments that are related to the overall membrane dimensions via the electron density profile. In particular, the location of minima in these form factors are correlated with the membrane dimensions as shown in Figs.~\ref{fig:quality} F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.

@@

%\subsection{Finding the best models for PC and PE lipid mixtures}
The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls}, showing areas per lipid from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. Because the area per lipid correlates with the average order parameter of the $\it{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1st for POPC with a clear advantage to the others, and 2nd for POPE with only marginally lower quality score than GROMOS-CKP and is therefore the best avilable choice for studies where packing effects of PE lipids are important. 
%The direct comparison with experiments in Fig.~\ref{fig:POPC_POPE_dataSI} in the supplementary information reveals that 
Simulations with CHARMM36 and GROMOS-CKP predict too packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while area per lipid for POPC is overstimated in Lipid17. For POPE, 
%order parameters in  Slipids and GROMOS-CKP are close to experiments suggesting that their are per lipids are most reasonable, while
CHARMM36 and Lipid17 predict too packed membranes.

In conclusion, 
%the quality evaluation based on the NMRlipids databank suggests that the Slipids parameters would give the most reliable description for membrane packing of POPC:POPE mixtures. Slipids was a relatively good model also for POPC:POPS mixtures in similar comparisons utilizing the preliminary version of the NMRlipids databank~\cite{antila22b}, although charged membranes are more complicated as the counterion binding affinity plays a significant role in membrane packing~\cite{antila19}. On the other hand, CHARMM36 overcomes Slipids when the focus is on conformational ensembles of headgroups  in different types of lipids because the accuracy of glycerol backbone is low in Slipids~\cite{bacle21}. Altogether, these examples demonstrate how 
the automatic quality evaluation in the NMRlipids databank enables rapid selection of the best simulation parameters to specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development and parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.

@@

\subsection{Detection rare phenomena using NMRlipids databank: Cholesterol flip-flops}
Lipid flip-flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip-flops are slow, occuring with the timescales of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster, yet the reported timescales range between minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip-flop events have been published recently~\cite{gu19,javanainen19,baral20}. 
%However, systematic studies on how cholesterol flip-flop rate depends on membrane properties are still beyond the current computational resources available for a single research group. 
These studies report increase in cholesterol flip-flop rates with increasing unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but correlations between cholesterol flip-flop rates and membrane properties have not been systematically studied. Here we demonstrate that the large amount of simulation data available in the NMRlipids databank with programmatic access makes the analysis of these correlations accessible for all.

Flip-flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-sn-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-sn-glycerol) in the NMRlipids databank. Cholesterol flip-flop rates range between 0.001-1.6\,$\mu$s$^{-1}$ with the mean of 0.16\,$\mu$s$^{-1}$ and median of 0.07\,$\mu$s$^{-1}$. These values are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. Flip-flop rates of DCHOL, 0.2\,$\mu$s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,$\mu$s$^{-1}$ and 0.5\,$\mu$s$^{-1}$, respectively. Flip-flops were not observed for other lipids, giving the upper limits for PC lipid flip-flop rate of 9\,s$^{-1}$ and for ceramide (N-palmitoyl-D-erythro-sphingosine) of 0.002\,$\mu$s$^{-1}$. Therefore, data in the NMRlipids databank suggests that the lipid flip-flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tablefootnote}
\usepackage{xr}

@@

\title{NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all}
%NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all^
%Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

\author[1]{Anne Kiirikki}         %ORCID: ?
\author[2]{Hanne Antila}          %ORCID: 0000-0002-2474-5053
\author[2]{Lara Bort}             %ORCID: ?
\author[2]{Pavel Buslaev}         %ORCID: ?
%\author[2]{Matti Javanainen}      %ORCID: 0000-0003-4858-364X
\author[2]{Batuhan Kav}           %ORCID: ?
\author[2]{Markus S. Miettinen}   %ORCID: 0000-0002-3999-4722
\author[2]{...}                   %ORCID: ?
\author[1,*]{O. H. Samuli Ollila} %ORCID: 0000-0002- 8728-1006
%\author[1,2,+]{Christine Author}
%\author[2,+]{Derek Author}
\affil[1]{University of Helsinki, Institute of Biotechnology, Helsinki, Finland}
\affil[2]{Affiliation, department, city, postcode, country}

@@

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally because at biological conditions they are in disordered fluid state, and sample preparation can be highly complicated due to the rich phase behaviour and complex interactions between lipids and other biomolecules. Consequently, the connections between complex lipid interactions and biological functions remain poorly understood.
%why such variety of lipids is needed and how their. 
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. 
%While MD simulations of systems with few biomolecules continue having major contributions in our understanding on biology and drug design, 
Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of 
%now searching new avenues by building models not only 
individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. In such systems exhibiting exceedingly complex behaviour,
model inaccuracies become increasingly difficult to detect and multiplication of small errors may have major impact on the conclusions.

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, such best practices are yet to be defined for MD simulations. Importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10} and in membranes~\cite{newport19,espigares20,leston22}, as well as for nucleic acids~\cite{hospital16}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}) and arbitrary biomolecules~\cite{bekker20}. However,
%The relevance of quality evaluation of simulation trajectories in databanks regarding technical details of simulations and accuracy of the underlying physical description of the system (force field) has become evident~\cite{tai04,meyer10,hospital20} and such 
automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access are still rare. 
%However, straightforward quality comparisons between individual simulations or force fields within these databanks remain challenging. 
In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

@@

Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best models for specific applications and accelerates the development of simulation parameters and methodology. The programmatic access enables users to apply data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields, both in academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics, and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation studies. Wider adaptation of the NMRlipids Databank will open even more possibities. 
%The practical relevance of the NMRlipids databank is exemplified by automatic quality evaluation and ranking of large amount of MD simulation data,

While the NMRlipids Databank currently contains only lipid bilayer systems, it can be readily expanded to contain other molecules, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The powerful combination of the overlay databank structure and open collaboration can be applied also when building databases in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Therefore, the approach introduced here opens up possibilities to create databases enabling data-driven and machine learning applications in new fields.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Programmatic access to MD simulation data of membranes composed of most common biologically abundant lipids}

NMRlipids Databank is a community driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. Nearly thousand simulation trajectories with the total length approaching half a millisecond can be accessed programmatically using Python API or graphical user interface (\url{http://www.databank.nmrlipids.fi/}). 
%NMRlipids databank is the only databank of MD simulations with open programmatic access to its content enabling fully flexible analyses of its content. 
Following the FAIR principles~\cite{wilkinson16} and NMRlipids project protocol, not only the content and computer programs related to the databank but also the whole construction process of the databank are open access~\cite{botan15}. 
%The NMRlipids databank will be always open for data and other contributions.

Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these compose the majority of available trajectories, also mixtures up to five lipids are available. The distribution of available simulations containing a specific lipid is shown in Fig.~\ref{fig:overlay}C. This roughly resembles the biological abundancies of different lipid types, PC being the most common followed by cholesterol, PE, PS, PG, PI, and other lipids depending on organism and organelle~\cite{vanmeer08}. The force fields used in the simulations cover all the essential parameters used in lipid simulations, ranging from highly popular CHARMM36 parameters to united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.
%Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

The efficient upcycling of large MD simulation trajectories in the NMRlipids databank is enabled by the overlay structure illustrated in Fig.~\ref{fig:overlay}A. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability and permanent links to the data, such as digital object identifiers (DOIs) or uniform resource names (URNs). The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored in the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses described in sections below. More detailed description on the NMRlipids databank structure is available the supplementary information.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay2.pdf}
    \caption{a) Structure of an overlay databank. 
    %Raw data is available from a publicly accessible repository (layer 1).
    %The core of the databank contains information on the location of the raw data and indexing of molecular %compositions and simulation details described with universal naming conventions (layer 2).
    %The content of databank can be viewed and analysed using external programs that automatically read the information and data from layers 1 and 2 (layer 3). Examples of such applications are the GitHub repository generating the results presented in this work and the graphical access to the databank.
    A more detailed structure of the databank layer in the NMRlipids databank is illustrated in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or fewer simulations ('others') are listed on the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    }
    \label{fig:overlay}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding the accuracy of the used force field parameters is crucial when estimating the reliability and significance of MD simulation results. However, the lack of universal quality measures for MD simulations and complicated landscape of force field quality for membranes hamper the quality of simulations which has resulted to many controversial results~\cite{antila22b}. For example, previous quality evaluations of membrane simulations have concluded that CHARMM36 parameters give the best description for lipid headgroup conformational ensembles~\cite{bacle21}, GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}, and that OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%, thereby being a major obstacle in many applications of membrane MD simulations. 
To enable the navigation in this complex force field quality landscape, we define here quantitative quality measures that can be used to rapidly find the best available parameters for a specific application or to guide force field development. For detailed definitions of quality measures, see the supplementary information.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality3.pdf}
    \caption{ A) Best simulations ranked based on overall order parameter quality.
    B) Best simulations ranked based on the overall order parameter quality of POPE lipid. 
    C)--E) Evaluation against experimental (X-ray scattering  NMR order parameter) data exemplified for a simulation with the best overall order parameter quality (C), the best quality for POPE lipid (D), and the headgroup quality for POPE (E).
    F) Scatter plots and Pearson correlation coefficients for the membrane area per lipid, thickness, first minima of x-ray scattering form factor and average order parameter of the {\textit sn}-1 acyl chain extracted from the NMRlipids databank. 
    All correlation coefficients have p-value below 0.001. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

Ranking of all simulations with the highest scores based on evaluation against C--H bond order parameters from NMR experiments is shown in Fig.~\ref{fig:quality} A and ranking of best POPE lipid bilayer simulations in Fig.~\ref{fig:quality} B. Direct comparisons with experiments for selected simulations are shown in Figs.~\ref{fig:quality} C--D. C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, but they correlate well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality} F. This is complemented by comparing the absolute values of X-ray scattering form factors between simulations and experiments that are related to the overall membrane dimensions \textit{via} electron density profile. Particularly, the location of minima in these form factors are correlated with the membrane dimensions as shown in Figs.~\ref{fig:quality} F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.

@@

%\subsection{Finding the best models for PC and PE lipid mixtures}
The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls} showing area per lipids from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. Because the area per lipid correlates with the average order parameter of $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F), simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1\textsuperscript{st} for POPC with a clear difference to others, and 2\textsuperscript{nd} for POPE with only marginally lower quality than GROMOS-CKP, therefore being the best selection for studies where packing effects of PE lipids are relevant. 
%The direct comparison with experiments in Fig.~\ref{fig:POPC_POPE_dataSI} in the supplementary information reveals that 
Simulations with CHARMM36 and GROMOS-CKP predict too packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while area per lipid for POPC is overestimated in Lipid17. For POPE, 
%order parameters in  Slipids and GROMOS-CKP are close to experiments suggesting that their are per lipids are most reasonable, while
CHARMM36 and Lipid17 predict too packed membranes.

In summary, 
%the quality evaluation based on the NMRlipids databank suggests that the Slipids parameters would give the most reliable description for membrane packing of POPC:POPE mixtures. Slipids was a relatively good model also for POPC:POPS mixtures in similar comparisons utilizing the preliminary version of the NMRlipids databank~\cite{antila22b}, although charged membranes are more complicated as the counterion binding affinity plays a significant role in membrane packing~\cite{antila19}. On the other hand, CHARMM36 overcomes Slipids when the focus is on conformational ensembles of headgroups  in different types of lipids because the accuracy of glycerol backbone is low in Slipids~\cite{bacle21}. Altogether, these examples demonstrate how 
the automatic quality evaluation of the NMRlipids databank enables the rapid selection of the best simulation parameters to specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development, and streamline the parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.

@@

\subsection{Detecting rare phenomena using NMRlipids databank: Cholesterol flip--flops}
Lipid flip--flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip--flops are slow, occurring with the timescales of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster, yet the reported timescales range from minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip--flop events have been published recently~\cite{gu19,javanainen19,baral20}. 
%However, systematic studies on how cholesterol flip-flop rate depends on membrane properties are still beyond the current computational resources available for a single research group. 
These studies report increase in cholesterol flip--flop rates with increasing acyl chain unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but correlations between cholesterol flip--flop rates and membrane properties have not been systematically studied. Here we demonstrate that the large amount of simulation data available in the NMRlipids databank, along with programmatic access providing the option to flexibly implement varied queries of the data, enables anybody to perform such analyses.

Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol) in the NMRlipids databank. Cholesterol flip--flop rates range between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$. These values are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9\,s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Therefore, data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.
